---
title: "LSE Practitioners' Challenge"
author: "Group 3"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(reshape2) 
library(lubridate)
library(gridExtra)
library(zoo)
library(tseries) 
library(car) 
library(moments)
library(knitr)
library(moments)
library(knitr)
library(quantmod)
library(PerformanceAnalytics)
library(timeSeries)
library(roll)
library(MASS)
library(extraDistr)
library(rugarch)
library(QRM)
library(kableExtra)
library(ggplot2)
```

```{r loading and cleaning data,echo = FALSE}
#------------------------- Cleaning Data ---------------------------------------
# Function to Load and Process Data
load_data <- function(file_path, price_col) {
  data <- read.csv(file_path, header = TRUE)
  
  # Check if the first column is the date
  if ("Date" %in% colnames(data)) {
    data$Date <- ymd(data$Date) 
  } else {
    # If Date is the index, reset it
    data <- data.frame(Date = rownames(data), coredata(data))
    data$Date <- as.Date(data$Date)
  }
  
  data <- data[order(data$Date), ]
  
  # Extract relevant columns
  prices <- data[, c("Date", price_col)]
  colnames(prices) <- c("date", "price")
  
  # Compute Simple Returns
  n <- nrow(prices)
  simple_returns <- numeric(n)
  simple_returns[2:n] <- (prices$price[2:n] - prices$price[1:(n - 1)]) / prices$price[1:(n - 1)]
  simple_returns[1] <- NA

  # Compute Log Returns
  log_returns <- numeric(n)
  log_returns[2:n] <- log(1 + simple_returns[2:n])
  log_returns[1] <- NA

  # Add to DataFrame
  prices$simple_returns <- simple_returns
  prices$log_returns <- log_returns
  prices <- prices[-1, ]  # Remove first row with NA
  
  return(prices)
}

# Load Data for Different Indices
SNP <- load_data("/Users/harveymiller/Documents/LSE/Year_4/Studies/Practitioners' Challenge/S&P 500 Index.csv", "S.P.500.Index")
NASDAQ <- load_data("/Users/harveymiller/Documents/LSE/Year_4/Studies/Practitioners' Challenge/NASDAQ Composite Index.csv", "NASDAQ.Composite.Index")
VIX <- load_data("/Users/harveymiller/Documents/LSE/Year_4/Studies/Practitioners' Challenge/S&P 500 VIX.csv", "S.P.500.Volatility.Index..VIX.")
```

### S&P500

## S&P500 Price & Returns Plot
```{r, fig.width=20, fig.height=6, fig.align='center',echo=FALSE}
# Convert log returns to zoo object
Returns.ts <- zoo(SNP$log_returns, order.by = SNP$date)

# Set up plotting area (1 row, 2 columns)
par(mfrow = c(1, 2),    
    cex.main = 2,    
    cex.lab = 1.4,             
    cex.axis = 1.2)  

# Plot Price Series
plot(SNP$date, 
     SNP$price,   
     type = "l", 
     ylab = "Price", 
     xlab = "Date", 
     col = "blue", 
     main = "Price Series",
     xaxt = "n")
axis.Date(1, at = seq(min(SNP$date), max(SNP$date), by = "1 year"), format = "%Y")

# Plot Log Returns
plot(index(Returns.ts), 
     coredata(Returns.ts), 
     type = 'l',
     ylab = "Log Returns", 
     xlab = "Date", 
     col = "red", 
     main = "Log Returns",
     lwd = 1,
     xaxt = "n")
axis.Date(1, at = seq(min(SNP$date), max(SNP$date), by = "1 year"), format = "%Y")

```

## Summary Statitsics for S&P500

```{r,echo = FALSE}
# Compute summary statistics
summary_stats = c(
  mean(SNP$log_returns, na.rm = TRUE),
  sd(SNP$log_returns, na.rm = TRUE),
  min(SNP$log_returns, na.rm = TRUE),
  max(SNP$log_returns, na.rm = TRUE),
  skewness(SNP$log_returns, na.rm = TRUE),
  kurtosis(SNP$log_returns, na.rm = TRUE),
  jarque.bera.test(SNP$log_returns)$p.value,  # Extract p-value
  Box.test(SNP$log_returns, type = "Ljung-Box")$p.value,  
  Box.test(SNP$log_returns^2, type = "Ljung-Box")$p.value  
)

# Convert to data frame
summary_stats_df <- data.frame(
  Metric = c("Mean", "SD", "Min", "Max", "Skewness", "Kurtosis", 
             "Jarque-Bera p-value", "Box-Ljung p-value", "Box-Ljung (squared) p-value"),
  Value = summary_stats
)

# Format as markdown table
kable(summary_stats_df, format = "markdown", digits = 4)

# Identify extreme return dates
min_return <- signif(min(SNP$log_returns, na.rm = TRUE) * 100, 2)
max_return <- signif(max(SNP$log_returns, na.rm = TRUE) * 100, 2)

# Print results
cat("Minimum return:", min_return, "\n")
cat("Maximum return:", max_return, "\n")
cat("Skewness:", signif(skewness(SNP$log_returns, na.rm = TRUE), 2), "\n")
cat("Kurtosis:", signif(kurtosis(SNP$log_returns, na.rm = TRUE), 2), "\n")

acf(SNP$log_returns, main = "ACF of Log Returns", lag.max = 200)
acf(SNP$log_returns^2, main = "ACF of Log Returns^2", lag.max = 200)

```


## GARCH Model Fit and In-Sample Diagnostics for S&P500

```{r, echo = FALSE}
# Specify the GARCH(1,1) model
GARCH_1_1 <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(1, 0), include.mean = TRUE),  # AR(1) mean model
  distribution.model = "norm"
)

# Residuals Analysis
GARCH_1_1_fit <- ugarchfit(spec = GARCH_1_1, data = SNP$log_returns, solver = "hybrid")
plot(GARCH_1_1_fit, which = 9)  # Standardized residuals are assumed to be normal in the model but not the case!


# Extract standardized residuals
std_residuals <- residuals(GARCH_1_1_fit, standardize = TRUE)

# Normality test for further proof
jarque.bera.test(std_residuals) # residuals are not normal, which violates assumption in garch, evidence to fit tGARCH


# ACF and Box test of residuals
acf(std_residuals, main = "ACF of Standardized Residuals") # this is good, as no autocorrelation present
Box.test(std_residuals, lag = 10, type = "Ljung-Box")
Box.test(std_residuals^2, lag = 10, type = "Ljung-Box")  # A significant p-value indicates that the model has not fully captured autocorrelation or volatility clustering.

```

## Out Of Sample Diagnostics for S&P500 (backtesting VaR and forecasting returns)
```{r VaR Forecast S&P,echo=FALSE}
#--------------------- GARCH Model Specification ---------------------
garch_spec <- ugarchspec(
  variance.model = list(garchOrder = c(1,1)), 
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "norm"
)

# Initialize Variables
VaR <- data.frame(date = SNP$date, price = SNP$price, log_returns = SNP$log_returns, GARCH = NA,forecasted_returns = NA)
p <- 0.01  # 5% significance level
WE <- 250  # Rolling window size

#---------------------- Rolling GARCH Estimation ----------------------
for (t in (WE+1):nrow(VaR)) {
  t1 <- t - WE
  t2 <- t - 1
  
  # Rolling window data
  window=VaR$log_returns[t1:t2]
  fit=ugarchfit(spec=garch_spec,data=window,solver = "hybrid")
  
  # we now forecast the conditional variance (T+1)
  s2=coef(fit)[1] +
    coef(fit)[2] * tail(window,1)^2 +
    coef(fit)[3] *tail(fit@fit$var,1 )
  VaR$GARCH[t]=-qnorm(p,sd = sqrt(s2))
    
  cat("Processed:", t, "\n")  # Track progress
  
  # Forecasted log return: ŷ_{t+1} = σ_{t+1} * z_{t+1}, z ~ N(0,1)
  z_t1 <- rnorm(1)  # Draw a random standard normal shock
  forecasted_return <- sqrt(s2) * z_t1
  VaR$forecasted_returns[t] <- forecasted_return
}

```

```{r plot,echo=FALSE}
# Ensure date format is correct
VaR$date <- as.Date(VaR$date)

# Define range for plotting
start_idx <- WE + 1
end_idx <- nrow(VaR)

# Plot VaR (red) vs. Actual Returns (blue)
plot(VaR$date[start_idx:end_idx], -VaR$GARCH[start_idx:end_idx], 
     type = "l", 
     col = "red", 
     ylab = "Returns ($)", 
     xlab = "Date", 
     main = "VaR(0.01) vs Returns (GARCH Model)", 
     ylim =  c(-0.3,0.1),
     lwd = 1)

lines(VaR$date[start_idx:end_idx], VaR$log_returns[start_idx:end_idx], 
      col = "blue", 
      lwd = 1)

# Add legend
legend("bottomright", 
       legend = c("VaR (0.01%) - GARCH", "Log Returns"), 
       col = c("red", "blue"), 
       lwd = c(2, 2),
       cex = 1)


# Calculate error metrics
rmse <- sqrt(mean((VaR$log_returns[start_idx:end_idx] - VaR$forecasted_returns[start_idx:end_idx])^2))
cat("\nRMSE:", rmse, "\n")

# Hit rate
valid_idx <- which(!is.na(VaR$log_returns) & !is.na(VaR$forecasted_returns))
sign_match <- sign(VaR$forecasted_returns[valid_idx]) == sign(VaR$log_returns[valid_idx])
hit_rate <- mean(sign_match)  
cat("\nHit Rate:", round(hit_rate * 100, 2), "%\n")


plot(VaR$date[start_idx:end_idx],VaR$log_returns[start_idx:end_idx], type = "l", col = "blue", 
     main = "Actual vs. Forecasted Returns", xlab = "Date", ylab = "Returns")
lines(VaR$date[start_idx:end_idx], VaR$forecasted_returns[start_idx:end_idx], col = "red")  
legend("topright", legend = c("Actual Returns", "Forecasted Returns"), col = c("blue", "red"), lty = c(1,1))

```

```{r, echo = FALSE}
V_GARCH=-VaR[251:2265,4]- VaR[251:2265,3] #var-returns

V_GARCH[V_GARCH<0]=0  # checks each element of the vector V and if an element is less than 0, it is replaced with 0.
V_GARCH[V_GARCH>0]=1  # checks each element of the vector V and if an element is greater than 0, it is replaced with 1.

V_GARCH=sum(V_GARCH)/(2264*0.01) 
```

Significantly underestimates risk, as VR ratio is 2.12

```{r Coverage Test S&P, echo = FALSE}
ra <- VaR[251:dim(VaR)[1],3]
VaRa <- VaR[251:dim(VaR)[1],"GARCH"]
eta <- ra < - VaRa # returns true (1) if ra < - VaRa holds, false otherwise (0)
v1 <- sum(eta)
v0 <- length(eta) - v1
picap <- v1 / (v1 + v0)

a <- (1 - p)^v0 * p^v1 # likelihood of restricted model
b <- (1 - picap)^v0 * picap^v1 # likelihood of unrestricted model

LR <- 2 * (log(b / a))
LR

if (LR > qchisq(p = 1 - p, df = 1)) {
  print('null hypothesis H0 is rejected')
} else {
  print('We cannot reject the null')
}
```

Following likelihood ratio test, for VaR(0.01) we can conclude that there is sufficient evidence to reject the null hypothesis that violations are bernoulli iid with probability 0.01 - not good!

## tGARCH Model Fit and In-Sample Diagnostics for S&P500

```{r in sample analysis S&P tGARCH, echo =FALSE}

# Specify the tGARCH(1,1) model
tGARCH_1_1 <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "std"  # t-distributed residuals
)

# Fit the model to log returns
tGARCH_1_1_fit <- ugarchfit(spec = tGARCH_1_1, data = SNP$log_returns, solver = "hybrid")
plot(tGARCH_1_1_fit, which = 9)  # Standardized residuals are assumed to be normal in the model but not the case!


# Extract standardized residuals
std_residuals <- residuals(tGARCH_1_1_fit, standardize = TRUE)

# Normality test for further proof
jarque.bera.test(std_residuals) # residuals are not normal, which violates assumption in garch, evidence to fit tGARCH


# ACF and Box test of residuals
acf(std_residuals, main = "ACF of Standardized Residuals") # this is good, as no autocorrelation present
Box.test(std_residuals, lag = 10, type = "Ljung-Box")
Box.test(std_residuals^2, lag = 10, type = "Ljung-Box")  # A significant p-value indicates that the model has not fully captured autocorrelation or volatility clustering.

```
## tGARCH Model Out-Of-Sample Diagnostics for S&P500 (backtesting VaR and forecasting returns)

```{r out of sample S&P tGARCH, echo = FALSE}
# tGARCH Model Specification
tGARCH_spec <- ugarchspec(
  variance.model = list(garchOrder = c(1,1)), 
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "std"
)

# Initialize DataFrame for VaR
VaR_t <- data.frame(date = SNP$date, price = SNP$price, log_returns = SNP$log_returns, tGARCH = NA,forecasted_returns = NA)
p <- 0.01  # 1% significance level
WE <- 250  # Rolling window size

# Rolling tGARCH Estimation
for (t in (WE+1):nrow(VaR_t)) {
  t1 <- t - WE
  t2 <- t - 1

  # Rolling window data
  window <- VaR_t$log_returns[t1:t2]

  # Check for NAs
  if (any(is.na(window))) {
    cat("Skipping iteration", t, "- NA values detected in window.\n")
    next
  }

  # Fit GARCH model with error handling
  fit <- tryCatch(
    ugarchfit(spec = tGARCH_spec, data = window, solver = "hybrid", fit.control = list(scale = 1)),
    error = function(e) {
      cat("GARCH model failed at iteration", t, "-", e$message, "\n")
      return(NULL)
    }
  )
  if (is.null(fit)) next

  # Extract Coefficients Safely
  coefs <- tryCatch(coef(fit), error = function(e) NULL)
  if (is.null(coefs) || length(coefs) < 3) {
    cat("Skipping iteration", t, "- coefficient extraction failed.\n")
    next
  }

  # Forecasting conditional variance (T+1)
  s2 <- coefs[1] + coefs[2] * tail(window,1)^2 + coefs[3] * tail(fit@fit$var,1)

  # Extract shape parameter safely
  shape_param <- tryCatch(coefs["shape"], error = function(e) NA)
  if (is.na(shape_param) || shape_param <= 2) {
    cat("Skipping iteration", t, "- invalid shape parameter.\n")
    next
  }

  # Compute VaR
  VaR_t$tGARCH[t] <- -sqrt(s2) * qt(p, df = shape_param) * sqrt((shape_param-2)/(shape_param))
  cat("Processed:", t, "\n")
  
    # Forecasted log return: ŷ_{t+1} = σ_{t+1} * z_{t+1}, z ~ N(0,1)
  z_t1 <- rt(1,df = shape_param)/ sqrt(shape_param / (shape_param - 2)) 
  forecasted_return <- sqrt(s2) * z_t1
  VaR_t$forecasted_returns[t] <- forecasted_return
}

# Ensure date format is correct
VaR_t$date <- as.Date(VaR_t$date)

# Define range for plotting
start_idx <- WE + 1
end_idx <- nrow(VaR_t)

# Plot VaR (red) vs. Actual Returns (blue)
plot(VaR_t$date[start_idx:end_idx], -VaR_t$tGARCH[start_idx:end_idx], 
     type = "l", col = "red", 
     ylab = "Returns ($)", 
     xlab = "Date", 
     main = "VaR vs Returns (tGARCH Model)", 
     ylim = c(-0.3, 0.1),
     lwd = 1)

lines(VaR_t$date[start_idx:end_idx], VaR_t$log_returns[start_idx:end_idx], 
      col = "blue", lwd = 1)

# Add legend
legend("bottomright", 
       legend = c("VaR (1%) - tGARCH", "Log Returns"), 
       col = c("red", "blue"), 
       lwd = c(2, 2),
       cex = 1)


valid_idx <- which(!is.na(VaR_t$log_returns[start_idx:end_idx]) & !is.na(VaR_t$forecasted_returns[start_idx:end_idx]))
rmse <- sqrt(mean((VaR_t$log_returns[start_idx:end_idx][valid_idx] - VaR_t$forecasted_returns[start_idx:end_idx][valid_idx])^2))
cat("\nRMSE:", rmse, "\n")

sign_match <- sign(VaR_t$forecasted_returns[start_idx:end_idx][valid_idx]) == sign(VaR_t$log_returns[start_idx:end_idx][valid_idx])
hit_rate <- mean(sign_match)  
cat("\nHit Rate:", round(hit_rate * 100, 2), "%\n")

plot(VaR_t$date[start_idx:end_idx],VaR_t$log_returns[start_idx:end_idx], type = "l", col = "blue", 
     main = "Actual vs. Forecasted Returns", xlab = "Date", ylab = "Returns")
lines(VaR_t$date[start_idx:end_idx], VaR_t$forecasted_returns[start_idx:end_idx], col = "red")  
legend("topright", legend = c("Actual Returns", "Forecasted Returns"), col = c("blue", "red"), lty = c(1,1))

```

```{r}
# Extract relevant data
VaR_subset <- VaR_t[251:2265, c(3, 4)]  # Select log_returns (col 3) and tGARCH (col 4)
VaR_subset <- na.omit(VaR_subset)

# Compute V_tGARCH only on non-NA values
V_tGARCH <- -VaR_subset[, 2] - VaR_subset[, 1]  # VaR - returns
V_tGARCH[V_tGARCH < 0] <- 0  # If negative, replace with 0
V_tGARCH[V_tGARCH > 0] <- 1  # If positive, replace with 1
VR_tGARCH <- sum(V_tGARCH) / (nrow(VaR_subset) * 0.01)  

cat("Violation Ratio (VR_tGARCH):", VR_tGARCH, "\n")

# Coverage Test
ra_t <- VaR_t[251:nrow(VaR_t), 3]
VaRa_t <- VaR_t[251:nrow(VaR_t), "tGARCH"]

# Remove NAs
valid_idx <- !is.na(ra_t) & !is.na(VaRa_t)
ra_t <- ra_t[valid_idx]
VaRa_t <- VaRa_t[valid_idx]

# Indicator function
eta_t <- ra_t < -VaRa_t

# Compute v1_t, v0_t, and picap_t
v1_t <- sum(eta_t)
v0_t <- length(eta_t) - v1_t
picap_t <- v1_t / (v1_t + v0_t)


a_t <- (1 - p)^v0_t * p^v1_t  # Likelihood of restricted model
b_t <- (1 - picap_t)^v0_t * picap_t^v1_t  # Likelihood of unrestricted model

LR_t <- 2 * (log(b_t / a_t))
LR_t

if (LR_t > qchisq(p = 1 - p, df = 1)) {
  print("Null hypothesis H0 is rejected")
} else {
  print("We cannot reject the null")
}

```


violation ratio more acceptable, test statistic also much lower which is good. tGARCH is better model.

### VIX 

## VIX Price & Returns
```{r, fig.width=20, fig.height=6, fig.align='center',echo=FALSE}
Returns.ts <- zoo(VIX$log_returns, order.by = VIX$date)
par(mfrow = c(1, 2), cex.main = 2, cex.lab = 1.4, cex.axis = 1.2)
plot(VIX$date, VIX$price, type = "l", col = "blue", main = "VIX Price Series", xaxt = "n")
axis.Date(1, at = seq(min(VIX$date), max(VIX$date), by = "1 year"), format = "%Y")
plot(index(Returns.ts), coredata(Returns.ts), type = 'l', col = "red", main = "VIX Log Returns", lwd = 1, xaxt = "n")
axis.Date(1, at = seq(min(VIX$date), max(VIX$date), by = "1 year"), format = "%Y")
```

## Summary Statistics for VIX
```{r,echo = FALSE}
summary_stats_vix = c(
  mean(VIX$log_returns, na.rm = TRUE),
  sd(VIX$log_returns, na.rm = TRUE),
  skewness(VIX$log_returns, na.rm = TRUE),
  kurtosis(VIX$log_returns, na.rm = TRUE),
  jarque.bera.test(VIX$log_returns)$p.value,
  Box.test(VIX$log_returns, type = "Ljung-Box")$p.value
)
summary_stats_df_vix <- data.frame(
  Metric = c("Mean", "SD", "Skewness", "Kurtosis", "Jarque-Bera p-value", "Box-Ljung p-value"),
  Value = summary_stats_vix
)
kable(summary_stats_df_vix, format = "markdown", digits = 4)
```

## GARCH Model Fit for VIX and In-Sample Diagnostics
```{r, echo = FALSE}
# Specify the GARCH(1,1) model
GARCH_1_1_VIX <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "norm"  # Assuming normal distribution for residuals
)

# Fit the model to log returns
GARCH_1_1_fit_VIX <- ugarchfit(spec = GARCH_1_1_VIX, data = VIX$log_returns, solver = "hybrid")

plot(GARCH_1_1_fit_VIX, which = 9)  # Standardized residuals

# Extract standardized residuals
std_residuals <- residuals(GARCH_1_1_fit_VIX, standardize = TRUE)

# Normality test for further proof
jarque.bera.test(std_residuals) # residuals are not normal, which violates assumption in garch, evidence to fit tGARCH


# ACF and Box test of residuals
acf(std_residuals, main = "ACF of Standardized Residuals") # this is good, as no autocorrelation present
Box.test(std_residuals, lag = 10, type = "Ljung-Box")
Box.test(std_residuals^2, lag = 10, type = "Ljung-Box")  # A significant p-value indicates that the model has not fully captured autocorrelation or volatility clustering.

```

## Out-of-Sample Diagnostics for VIX (backtesting VaR and forecasting returns)
```{r VaR Forecast,echo=FALSE}
#--------------------- GARCH Model Specification ---------------------
garch_spec_VIX <- ugarchspec(
  variance.model = list(garchOrder = c(1,1)), 
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "norm"
)

# Initialize Variables
VaR_VIX <- data.frame(date = VIX$date, price = VIX$price, log_returns = VIX$log_returns, GARCH_VaR = NA,forecasted_returns = NA)
p <- 0.01  # 1% significance level
WE <- 250  # Rolling window size

#---------------------- Rolling GARCH Estimation ----------------------
for (t in (WE+1):nrow(VaR_VIX)) {
  t1 <- t - WE
  t2 <- t - 1
  
  # Rolling window data
  window_returns <- VaR_VIX$log_returns[t1:t2]
  
  # Fit GARCH(1,1)
  garch_fit_VIX <- ugarchfit(spec = garch_spec_VIX, data = window_returns, solver = 'hybrid')
  
  # we now forecast the conditional variance (T+1)
  s2=coef(garch_fit_VIX)[1] +
    coef(garch_fit_VIX)[2] * tail(window_returns,1)^2 +
    coef(garch_fit_VIX)[3] *tail(garch_fit_VIX@fit$var,1 )

  # Compute VaR
  VaR_VIX$GARCH_VaR[t] <- -qnorm(p,sd = sqrt(s2))  # VaR at 1%
  cat("Processed:", t, "\n")  # Track progress
  
  # Forecasted log return: ŷ_{t+1} = σ_{t+1} * z_{t+1}, z ~ N(0,1)
  z_t1 <- rnorm(1)  # Draw a random standard normal shock
  forecasted_return <- sqrt(s2) * z_t1
  VaR_VIX$forecasted_returns[t] <- forecasted_return
}


```

```{r,echo=FALSE}
# Ensure date format is correct
VaR_VIX$date <- as.Date(VaR_VIX$date)

# Define range for plotting
start_idx <- WE + 1
end_idx <- nrow(VaR_VIX)

# Plot VaR (red) vs. Actual Returns (blue)
plot(VaR_VIX$date[start_idx:end_idx], -VaR_VIX$GARCH_VaR[start_idx:end_idx], 
     type = "l", 
     col = "red", 
     ylab = "Returns ($)", 
     xlab = "Date", 
     main = "VaR vs Returns (GARCH Model for VIX)", 
     ylim = c(-1.3,1.5),
     lwd = 1)

lines(VaR_VIX$date[start_idx:end_idx], VaR_VIX$log_returns[start_idx:end_idx], 
      col = "blue", 
      lwd = 1)

# Add legend
legend("bottomright", 
       legend = c("VaR (0.01%) - GARCH", "Log Returns"), 
       col = c("red", "blue"), 
       lwd = c(2, 2),
       cex = 1)


# Calculate error metrics
rmse <- sqrt(mean((VaR_VIX$log_returns[start_idx:end_idx] - VaR_VIX$forecasted_returns[start_idx:end_idx])^2))
cat("\nRMSE:", rmse, "\n")

valid_idx <- which(!is.na(VaR_VIX$log_returns[start_idx:end_idx]) & !is.na(VaR_VIX$forecasted_returns[start_idx:end_idx]))
sign_match <- sign(VaR_VIX$forecasted_returns[start_idx:end_idx][valid_idx]) == sign(VaR_VIX$log_returns[start_idx:end_idx][valid_idx])
hit_rate <- mean(sign_match)  
cat("\nHit Rate:", round(hit_rate * 100, 2), "%\n")

plot(VaR_VIX$date[start_idx:end_idx],VaR_VIX$log_returns[start_idx:end_idx], type = "l", col = "blue", 
     main = "Actual vs. Forecasted Returns", xlab = "Date", ylab = "Returns",ylim = c(-1,1))
lines(VaR_VIX$date[start_idx:end_idx], VaR_VIX$forecasted_returns[start_idx:end_idx], col = "red")  
legend("bottomright", legend = c("Actual Returns", "Forecasted Returns"), col = c("blue", "red"), lty = c(1,1))

```

```{r, echo = FALSE}
V_GARCH_VIX = -VaR_VIX[251:2265,4] - VaR_VIX[251:2265,3] #var-returns

V_GARCH_VIX[V_GARCH_VIX<0] = 0  # If element < 0, replace with 0
V_GARCH_VIX[V_GARCH_VIX>0] = 1  # If element > 0, replace with 1

V_GARCH_VIX = sum(V_GARCH_VIX)/(2264*0.01) 
V_GARCH_VIX

# Coverage Test
ra_t <- VaR_VIX[251:nrow(VaR_t), 3]
VaRa_t <- VaR_VIX[251:nrow(VaR_t), "GARCH_VaR"]

# Remove NAs
valid_idx <- !is.na(ra_t) & !is.na(VaRa_t)
ra_t <- ra_t[valid_idx]
VaRa_t <- VaRa_t[valid_idx]

# Indicator function
eta_t <- ra_t < -VaRa_t

# Compute v1_t, v0_t, and picap_t
v1_t <- sum(eta_t)
v0_t <- length(eta_t) - v1_t
picap_t <- v1_t / (v1_t + v0_t)


a_t <- (1 - p)^v0_t * p^v1_t  # Likelihood of restricted model
b_t <- (1 - picap_t)^v0_t * picap_t^v1_t  # Likelihood of unrestricted model

LR_t <- 2 * (log(b_t / a_t))
LR_t

if (LR_t > qchisq(p = 1 - p, df = 1)) {
  print("Null hypothesis H0 is rejected")
} else {
  print("We cannot reject the null")
}

```


### NASDAQ 

## NASDAQ Price & Returns
```{r, fig.width=20, fig.height=6, fig.align='center',echo=FALSE}
Returns.ts <- zoo(NASDAQ$log_returns, order.by = NASDAQ$date)
par(mfrow = c(1, 2), cex.main = 2, cex.lab = 1.4, cex.axis = 1.2)
plot(NASDAQ$date, NASDAQ$price, type = "l", col = "blue", main = "NASDAQ Price Series", xaxt = "n")
axis.Date(1, at = seq(min(NASDAQ$date), max(NASDAQ$date), by = "1 year"), format = "%Y")
plot(index(Returns.ts), coredata(Returns.ts), type = 'l', col = "red", main = "NASDAQ Log Returns", lwd = 1, xaxt = "n")
axis.Date(1, at = seq(min(NASDAQ$date), max(NASDAQ$date), by = "1 year"), format = "%Y")
```
## Summary Statistics for NASDAQ

```{r, echo = FALSE}
summary_stats_nasdaq = c(
  mean(NASDAQ$log_returns, na.rm = TRUE),
  sd(NASDAQ$log_returns, na.rm = TRUE),
  skewness(NASDAQ$log_returns, na.rm = TRUE),
  kurtosis(NASDAQ$log_returns, na.rm = TRUE),
  jarque.bera.test(NASDAQ$log_returns)$p.value,
  Box.test(NASDAQ$log_returns, type = "Ljung-Box")$p.value
)
summary_stats_df_nasdaq <- data.frame(
  Metric = c("Mean", "SD", "Skewness", "Kurtosis", "Jarque-Bera p-value", "Box-Ljung p-value"),
  Value = summary_stats_nasdaq
)
kable(summary_stats_df_nasdaq, format = "markdown", digits = 4)

```
## GARCH Model Fit and In-Sample Diagnostics for NASDAQ

```{r, echo = FALSE}
# Specify the GARCH(1,1) model
GARCH_1_1_NASDAQ <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "norm"
)

# Fit the model to log returns
GARCH_1_1_fit_NASDAQ <- ugarchfit(spec = GARCH_1_1_NASDAQ, data = NASDAQ$log_returns, solver = "hybrid")

plot(GARCH_1_1_fit_NASDAQ, which = 9)  # Standardized residuals

# Extract standardized residuals
std_residuals <- residuals(GARCH_1_1_fit_NASDAQ, standardize = TRUE)

# Normality test for further proof
jarque.bera.test(std_residuals) # residuals are not normal, which violates assumption in garch, evidence to fit tGARCH


# ACF and Box test of residuals
acf(std_residuals, main = "ACF of Standardized Residuals") # this is good, as no autocorrelation present
Box.test(std_residuals, lag = 10, type = "Ljung-Box")
Box.test(std_residuals^2, lag = 10, type = "Ljung-Box")  # A significant p-value indicates that the model has not fully captured autocorrelation or volatility clustering.

```
## Out Of Sample Diagnostics for NASDAQ (backtesting VaR and forecasting returns)

```{r, echo = FALSE}
#--------------------- GARCH Model Specification ---------------------
garch_spec_NASDAQ <- ugarchspec(
  variance.model = list(garchOrder = c(1,1)), 
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "norm"
)

# Initialize Variables
VaR_NASDAQ <- data.frame(date = NASDAQ$date, price = NASDAQ$price, log_returns = NASDAQ$log_returns, GARCH_VaR = NA,forecasted_returns = NA)
p <- 0.01  # 1% significance level
WE <- 250  # Rolling window size

#---------------------- Rolling GARCH Estimation ----------------------
for (t in (WE+1):nrow(VaR_NASDAQ)) {
  t1 <- t - WE
  t2 <- t - 1
  
  # Rolling window data
  window_returns <- VaR_NASDAQ$log_returns[t1:t2]
  
  # Fit GARCH(1,1)
  garch_fit_NASDAQ <- ugarchfit(spec = garch_spec_NASDAQ, data = window_returns, solver = 'hybrid')
  
  # we now forecast the conditional variance (T+1)
  s2=coef(garch_fit_NASDAQ)[1] +
    coef(garch_fit_NASDAQ)[2] * tail(window_returns,1)^2 +
    coef(garch_fit_NASDAQ)[3] *tail(garch_fit_NASDAQ@fit$var,1 )

  # Compute VaR
  VaR_NASDAQ$GARCH_VaR[t] <- -qnorm(p,sd = sqrt(s2))  # VaR at 1%
  cat("Processed:", t, "\n")  # Track progress
  
  # Forecasted log return: ŷ_{t+1} = σ_{t+1} * z_{t+1}, z ~ N(0,1)
  z_t1 <- rnorm(1)  # Draw a random standard normal shock
  forecasted_return <- sqrt(s2) * z_t1
  VaR_NASDAQ$forecasted_returns[t] <- forecasted_return
}

```


```{r, echo = FASLE}
# Ensure date format is correct
VaR_NASDAQ$date <- as.Date(VaR_NASDAQ$date)

# Define range for plotting
start_idx <- WE + 1
end_idx <- nrow(VaR_NASDAQ)

# Plot VaR (red) vs. Actual Returns (blue)
plot(VaR_NASDAQ$date[start_idx:end_idx], -VaR_NASDAQ$GARCH_VaR[start_idx:end_idx], 
     type = "l", 
     col = "red", 
     ylab = "Returns ($)", 
     xlab = "Date", 
     main = "VaR vs Returns (GARCH Model for NASDAQ)", 
     ylim = c(-0.3,0.3),
     lwd = 1)

lines(VaR_NASDAQ$date[start_idx:end_idx], VaR_NASDAQ$log_returns[start_idx:end_idx], 
      col = "blue", 
      lwd = 1)

# Add legend
legend("bottomright", 
       legend = c("VaR (0.01%) - GARCH", "Log Returns"), 
       col = c("red", "blue"), 
       lwd = c(2, 2),
       cex = 1)

# Calculate error metrics
rmse <- sqrt(mean((VaR_NASDAQ$log_returns[start_idx:end_idx] - VaR_NASDAQ$forecasted_returns[start_idx:end_idx])^2))
cat("\nRMSE:", rmse, "\n")

valid_idx <- which(!is.na(VaR_NASDAQ$log_returns[start_idx:end_idx]) & !is.na(VaR_NASDAQ$forecasted_returns[start_idx:end_idx]))
sign_match <- sign(VaR_NASDAQ$forecasted_returns[start_idx:end_idx][valid_idx]) == sign(VaR_NASDAQ$log_returns[start_idx:end_idx][valid_idx])
hit_rate <- mean(sign_match)  
cat("\nHit Rate:", round(hit_rate * 100, 2), "%\n")

plot(VaR_NASDAQ$date[start_idx:end_idx],VaR_NASDAQ$log_returns[start_idx:end_idx], type = "l", col = "blue", 
     main = "Actual vs. Forecasted Returns", xlab = "Date", ylab = "Returns")
lines(VaR_NASDAQ$date[start_idx:end_idx], VaR_NASDAQ$forecasted_returns[start_idx:end_idx], col = "red")  
legend("bottomright", legend = c("Actual Returns", "Forecasted Returns"), col = c("blue", "red"), lty = c(1,1))


```

```{r, echo = FALSE}
V_GARCH_NASDAQ = -VaR_NASDAQ[251:2265,4] - VaR_NASDAQ[251:2265,3] #var-returns

V_GARCH_NASDAQ[V_GARCH_NASDAQ<0] = 0  # If element < 0, replace with 0
V_GARCH_NASDAQ[V_GARCH_NASDAQ>0] = 1  # If element > 0, replace with 1

V_GARCH_NASDAQ = sum(V_GARCH_NASDAQ)/(2264*0.01) 
V_GARCH_NASDAQ

ra <- VaR_NASDAQ[251:dim(VaR_NASDAQ)[1],3]
VaRa <- VaR_NASDAQ[251:dim(VaR_NASDAQ)[1],"GARCH_VaR"]
eta <- ra < - VaRa # returns true (1) if ra < - VaRa holds, false otherwise (0)
v1 <- sum(eta)
v0 <- length(eta) - v1
picap <- v1 / (v1 + v0)

a <- (1 - p)^v0 * p^v1 # likelihood of restricted model
b <- (1 - picap)^v0 * picap^v1 # likelihood of unrestricted model

LR <- 2 * (log(b / a))
LR

if (LR > qchisq(p = 1 - p, df = 1)) {
  print('null hypothesis H0 is rejected')
} else {
  print('We cannot reject the null')
}
```

## tGARCH Model Fit and In-Sample Diagnostics for NASDAQ

```{r in sample analysis, echo =FALSE}

# Specify the tGARCH(1,1) model
tGARCH_1_1 <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "std"  # t-distributed residuals
)

# Fit the model to log returns
tGARCH_1_1_fit <- ugarchfit(spec = tGARCH_1_1, data = NASDAQ$log_returns, solver = "hybrid")

# Plot conditional standard deviation (volatility)
plot(tGARCH_1_1_fit, which = 3)

# Plot standardized residuals
plot(tGARCH_1_1_fit, which = 9)

std_residuals <- residuals(tGARCH_1_1_fit, standardize = TRUE)

# Normality test for further proof
jarque.bera.test(std_residuals) # residuals are not normal, which violates assumption in garch, evidence to fit tGARCH


# ACF and Box test of residuals
acf(std_residuals, main = "ACF of Standardized Residuals") # this is good, as no autocorrelation present
Box.test(std_residuals, lag = 10, type = "Ljung-Box")
Box.test(std_residuals^2, lag = 10, type = "Ljung-Box")  # A significant p-value indicates that the model has not fully captured autocorrelation or volatility clustering.

```
## tGARCH Model Out-Of-Sample Diagnostics for NASDAQ (backtesting VaR and forecasting returns)

```{r out of sample, echo = FALSE}
# tGARCH Model Specification
tGARCH_spec <- ugarchspec(
  variance.model = list(garchOrder = c(1,1)), 
  mean.model = list(armaOrder = c(0,0), include.mean = FALSE),
  distribution.model = "std"
)

# Initialize DataFrame for VaR
VaR_t_NASDAQ <- data.frame(date = NASDAQ$date, price = NASDAQ$price, log_returns = NASDAQ$log_returns, tGARCH = NA, forecasted_returns = NA)
p <- 0.01  # 1% significance level
WE <- 250  # Rolling window size

# Rolling tGARCH Estimation
for (t in (WE+1):nrow(VaR_t_NASDAQ)) {
  t1 <- t - WE
  t2 <- t - 1

  # Rolling window data
  window <- VaR_t_NASDAQ$log_returns[t1:t2]

  # Check for NAs
  if (any(is.na(window))) {
    cat("Skipping iteration", t, "- NA values detected in window.\n")
    next
  }

  # Fit GARCH model with error handling
  fit <- tryCatch(
    ugarchfit(spec = tGARCH_spec, data = window, solver = "hybrid", fit.control = list(scale = 1)),
    error = function(e) {
      cat("GARCH model failed at iteration", t, "-", e$message, "\n")
      return(NULL)
    }
  )
  if (is.null(fit)) next

  # Extract Coefficients Safely
  coefs <- tryCatch(coef(fit), error = function(e) NULL)
  if (is.null(coefs) || length(coefs) < 3) {
    cat("Skipping iteration", t, "- coefficient extraction failed.\n")
    next
  }

  # Forecasting conditional variance (T+1)
  s2 <- coefs[1] + coefs[2] * tail(window,1)^2 + coefs[3] * tail(fit@fit$var,1)

  # Extract shape parameter safely
  shape_param <- tryCatch(coefs["shape"], error = function(e) NA)
  if (is.na(shape_param) || shape_param <= 2) {
    cat("Skipping iteration", t, "- invalid shape parameter.\n")
    next
  }

  # Compute VaR
  VaR_t_NASDAQ$tGARCH[t] <- -sqrt(s2) * qt(p, df = shape_param) * sqrt((shape_param-2)/(shape_param))
  cat("Processed:", t, "\n")
  
  # Forecasted log return: ŷ_{t+1} = σ_{t+1} * z_{t+1}, z ~ N(0,1)
  z_t1 <- rt(1,df = shape_param)/ sqrt(shape_param / (shape_param - 2)) 
  forecasted_return <- sqrt(s2) * z_t1
  VaR_t_NASDAQ$forecasted_returns[t] <- forecasted_return
}

# Ensure date format is correct
VaR_t_NASDAQ$date <- as.Date(VaR_t_NASDAQ$date)

# Define range for plotting
start_idx <- WE + 1
end_idx <- nrow(VaR_t_NASDAQ)

# Plot VaR (red) vs. Actual Returns (blue)
plot(VaR_t_NASDAQ$date[start_idx:end_idx], -VaR_t_NASDAQ$tGARCH[start_idx:end_idx], 
     type = "l", col = "red", 
     ylab = "Returns ($)", 
     xlab = "Date", 
     main = "VaR vs Returns (tGARCH Model)", 
     ylim = c(-0.3, 0.1),
     lwd = 1)

lines(VaR_t_NASDAQ$date[start_idx:end_idx], VaR_t_NASDAQ$log_returns[start_idx:end_idx], 
      col = "blue", lwd = 1)

# Add legend
legend("bottomright", 
       legend = c("VaR (1%) - tGARCH", "Log Returns"), 
       col = c("red", "blue"), 
       lwd = c(2, 2),
       cex = 1)

valid_idx <- which(!is.na(VaR_t_NASDAQ$log_returns[start_idx:end_idx]) & !is.na(VaR_t_NASDAQ$forecasted_returns[start_idx:end_idx]))
rmse <- sqrt(mean((VaR_t_NASDAQ$log_returns[start_idx:end_idx][valid_idx] - VaR_t_NASDAQ$forecasted_returns[start_idx:end_idx][valid_idx])^2))
cat("\nRMSE:", rmse, "\n")

valid_idx <- which(!is.na(VaR_t_NASDAQ$log_returns[start_idx:end_idx]) & !is.na(VaR_t_NASDAQ$forecasted_returns[start_idx:end_idx]))
sign_match <- sign(VaR_t_NASDAQ$forecasted_returns[start_idx:end_idx][valid_idx]) == sign(VaR_t_NASDAQ$log_returns[start_idx:end_idx][valid_idx])
hit_rate <- mean(sign_match)  
cat("\nHit Rate:", round(hit_rate * 100, 2), "%\n")

plot(VaR_t_NASDAQ$date[start_idx:end_idx],VaR_t_NASDAQ$log_returns[start_idx:end_idx], type = "l", col = "blue", 
     main = "Actual vs. Forecasted Returns", xlab = "Date", ylab = "Returns")
lines(VaR_t_NASDAQ$date[start_idx:end_idx], VaR_t_NASDAQ$forecasted_returns[start_idx:end_idx], col = "red")  
legend("bottomright", legend = c("Actual Returns", "Forecasted Returns"), col = c("blue", "red"), lty = c(1,1))
```


```{r}
# Extract relevant data
VaR_subset <- VaR_t_NASDAQ[251:2265, c(3, 4)]  # Select log_returns (col 3) and tGARCH (col 4)
VaR_subset <- na.omit(VaR_subset)

# Compute V_tGARCH only on non-NA values
V_tGARCH <- -VaR_subset[, 2] - VaR_subset[, 1]  # VaR - returns
V_tGARCH[V_tGARCH < 0] <- 0  # If negative, replace with 0
V_tGARCH[V_tGARCH > 0] <- 1  # If positive, replace with 1
VR_tGARCH <- sum(V_tGARCH) / (nrow(VaR_subset) * 0.01)  

cat("Violation Ratio (VR_tGARCH):", VR_tGARCH, "\n")

# Coverage Test
ra_t <- VaR_t_NASDAQ[251:nrow(VaR_t), 3]
VaRa_t <- VaR_t_NASDAQ[251:nrow(VaR_t), "tGARCH"]

# Remove NAs
valid_idx <- !is.na(ra_t) & !is.na(VaRa_t)
ra_t <- ra_t[valid_idx]
VaRa_t <- VaRa_t[valid_idx]

# Indicator function
eta_t <- ra_t < -VaRa_t

# Compute v1_t, v0_t, and picap_t
v1_t <- sum(eta_t)
v0_t <- length(eta_t) - v1_t
picap_t <- v1_t / (v1_t + v0_t)


a_t <- (1 - p)^v0_t * p^v1_t  # Likelihood of restricted model
b_t <- (1 - picap_t)^v0_t * picap_t^v1_t  # Likelihood of unrestricted model

LR_t <- 2 * (log(b_t / a_t))
LR_t

if (LR_t > qchisq(p = 1 - p, df = 1)) {
  print("Null hypothesis H0 is rejected")
} else {
  print("We cannot reject the null")
}

```

